### 2011-11-07 doc/pldi2012/highlevel.tex
Some minor fixes to the high level section

%When performing dependence analysis the runtime relies on its knowledge about
%the relationships between logical regions and partitions.  We first describe
%the the components to dependence analysis in sections \ref{subsec:regiontree},
%\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
%dependence analysis on this data structures in section \ref{subsec:depdetect}.
%\subsubsection{Region Trees}
%\label{subsec:regiontree}
%To detect dependences between the tasks, the runtime leverages its knowledge
%about logical regions and their partitions.  To create a new logical region
%or partition, the application must invoke the corresponding call in the
%runtime.  As these calls occur, the runtime constructs a data strcture
%called a {\em region tree}.  A region tree describes the relationship between
%logical regions and partitions.  A region tree contains two types of nodes:
%region nodes and partition nodes.  Every region tree is rooted with a region
%node and alternates between region and partition nodes at each corresponding level.
%A region node represents a specific logical region and tracks the set of
%partitions of that logical region.  Similarly a partition node represents 
%a specific partitioning of a logical region and keeps track of the logical regions 
%which are subregions of the partition.  
\subsubsection{Dependencies From Coherence and Privileges}
\label{subsec:cohdep}
Conceptually, dependence detection between a new task $t$ and a previous task $t'$ requires comparing the two sets of
logical regions accessed.  For each logical region from $t'$
that may overlap with a logical region needed by $t$, the privileges
and coherence modes are compared to determine whether a dependence exists.  Read
privileges never cause conflicts, but if either 
task needs write or reduction privileges, the coherence modes are compared:

@@

To detect dependencies efficiently, the high-level runtime maintains a
local region forest (recall Section~\ref{sec:ex}) for each task.  The
root regions are the task's region arguments, and the forest includes
all partitions and subregions created within the task.  To perform
dependence detection on a region $r$, the runtime first finds a root
region $r'$ such that $r \rleq r'$.  The runtime then walks the region
forest from $r'$ to $r$.  Each region $r''$ on the path contains a
list $L$ of registered tasks using $r''$, and potential dependencies
are discovered by comparing privileges and coherence modes with the
tasks in $L$.

The runtime next checks if a partition of $r''$ is {\em open}.  An
open partition indicates that there are active tasks using subregions
of $r''$.  No more than one partition of a node can be open
at any time.  If the open partition is is along the path to $r$,
traversal continues.  If the open partition lies along another 
path the runtime must {\em close} the currently active partition because it cannot guarantee
disjointness with the region the new task wishes to access.  To close a
partition, the runtime finds all active tasks in that
subtree and considers them all dependencies for the new task.  
With the previously active partition closed, the runtime can 
then opens the desired partition and traverses further down the tree as
needed.

@@

A {\em physical instance} is a copy of the data in a logical region.  Multiple
physical instances of a logical region may exist simultaneously, and it is the responsibility of the high-level runtime
to ensure tasks use the correct version.
In an out-of-order processor this problem is solved by register-renaming which maps logical
registers to a larger set of physical registers.  Our solution employs a similar technique
that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
is a place holder for the set of physical instances a task will reference when it actually runs.  
Abstract instances are needed because we may not know the actual physical instances a task will
use until its predecessors in the dependence graph are {\em mapped} 
(see Section \ref{subsec:mapping}).  
To determine the abstract instances of a region, we keep track of 
the previous valid abstract instances that were observed while traversing the region forest
during dependence analysis.  Every region in the region forest that has active tasks
has a valid abstract instance.  Every time we close a region we close all the valid
abstract instances.  When a region is reopened it must then create a new abstract instance
to reflect that new physical instances must be created (i.e., a fresh version of the data must be copied from
the parent region).  In addition, whenever a task writes a region, a new abstract instance is usually created
to represent the new version of the data.

@@

For the ghost node regions, dependence analysis discovers 
a dependence between each of the {\tt distribute\_charge} tasks and all of the
{\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
performing writes and the partition is aliased.  However, the runtime will detect this as
anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
create new valid abstract instances rather than use the same abstract instances as
each of the {\tt compute\_new\_current} tasks.  There are no dependences detected
between each of the {\tt distribute\_charge} tasks on the ghost regions because
the {\tt Simultaneous} coherence property permits multiple tasks to be using the same
partition at the same time even if the subregions they are accessing are aliased.  The
state of the region forest after dependence analysis for the first {\tt distribute\_charge} task can
be seen in figure~\ref{sfig:mapping_fig:dc}.

@@

\subsection{Mapping}
\label{subsec:mapping}
After dependence analysis a task is placed on either a {\em ready queue} to be mapped or a 
The high-level runtime maintains the invariant that a task {\em maps} only after all tasks on which it depends have mapped. 
While some predecessor tasks have not mapped a task remains in the {\em waiting queue}; otherwise it is placed on the {\em ready queue}.

@@

These eight mapper calls give the programmer complete control over where all tasks are executed, and
where all data is placed within the memory hierarchy.  While this interface allows the programmer
to tune performance, it cannot affect correctness.  The high-level runtime still manages all the task dependences and the
ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
made by any mapper, the resulting program will always be executed correctly.
To make it easy for programmers to write programs, we also provide a default
 mapper that uses simple heuristics to place tasks and data.  
%For example, when
%performing task stealing, the default mapper only encourages stealing tasks in groups that use
%the same logical regions, which improves locality by moving tasks that use the same regions
%together.  The default mapper implementation makes it easy for the programmer to get an 
%initial program running.  As the programmer tunes the application, he can write mappers that
%specialize for both application specific and/or machine specific parameters.

The high-level runtime also supports the use of multiple mappers in the same application.  Each task 
call may (but does not need to) have an associated mapper,
%runtime  can choose a mapper to be used for the runtime calls made inside that
%task.  
which allows for the creation of library code that provides its own mappers ind pendent
from any application level mappers.

@@

When a processor notices that it has no work on its ready queue, it will inquire into each
mapper asking for a target processor to attempt a steal.  A mapper can either respond with
a target processor, or indicate that no steal should be attempted.  The runtime will then
send steal requests to each processor on behalf of each specific mapper.

When a steal request is received at a processor it will determine the mapper(s) that generated
the list of steal requests.  For each mapper that issued a steal request, the runtime will
give the local instance of that mapper a list of the tasks in the ready queue that that
specific mapper is responsible for mapping.  The mapper must then determine the set which
it is going to permit to be stolen, if any.  Any tasks that the mapper marks as stolen are
sent back to the stealing processor.  By only allowing mappers to make steal requests about
the tasks they own, we give mappers complete control over where tasks are mapped.
In addition, we also give the mappers control over how steal requests are made for
the case where the mapper has better knowledge about where additional work is than the 
runtime.  Note that this mapping interface is sufficiently powerful to allow a user to
disable all stealing if desired, and to map tasks to specific processors with no interference
from the runtime.

\subsection{Execution}
\label{subsec:execution}
Once task mapping is complete the high-level runtime performs the necessary operations
to issue the task to the low level runtime.  It first creates any new physical instances
for the regions the task requires.  It then issues copy operations required
for the task to execute contingent on predecessor tasks producing the source physical instances
having finished.  Then it issues the task itself to the low level processor contingent
on all the copies for the task having finished as well as the the set of tasks
on which the issuing task depends having finished.  These dependences are all encoded
using {\em events} (see Section \ref{subsec:events}).  Continuing the analogy of an out-of-order
processor, the execution phase is equivalent to issuing an instruction to an
execution unit, and having the execution unit defer the instruction's execution until all its
operands are ready.  The low-level runtime contains processors that act as execution
units that will run a task once all of the events required for the task to run
have occurred.

@@

\subsection{Task Completion}
\label{subsec:cleanup}
We define a task to have {\tt completed} once all of the task's regions reflect
the updated state of the task and all its sub-tasks.  
%This reflects an underlying assumption 
%of the mapping process that the regions for which a task has priviliges will represent the data produced 
%by that task.  However, this is not inherently true that a physical instance of a task reflects
%this state after the task finishes executing.  
It is possible for a task to terminate but to have subtasks that are still in the execution phase.
%If a task has child tasks these tasks may
%still be executing when the parent task finishes the execution phase.  
To handle this scenario the runtime must compute the necessary copy operations
from child task physical instances back to the parent task's physical instances.  To
compute these copy operations, the runtime must first wait until all the child tasks
have been mapped.  The runtime then issues a close operation to the each of the parent
task's region forests.  These close operations record all of the necessary copies to
restore the parent task's regions.  
%We call these copy operations at the end of  a task {\em restoring copies}.
The runtime issues these copies with the required dependences
and marks the task completed once the copy operations are complete.
This is similar to the write-back phase of an out-of-order processor where store instructions cannot be
considered complete until the store effects the state of memory.

>>>>

%When performing dependence analysis the runtime relies on its knowledge about
%the relationships between logical regions and partitions.  We first describe
%the the components to dependence analysis in sections \ref{subsec:regiontree},
%\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
%dependence analysis on this data structures in section \ref{subsec:depdetect}.
%\subsubsection{Region Trees}
%\label{subsec:regiontree}
%To detect dependences between the tasks, the runtime leverages its knowledge
%about logical regions and their partitions.  To create a new logical region
%or partition, the application must invoke the corresponding call in the
%runtime.  As these calls occur, the runtime constructs a data strcture
%called a {\em region tree}.  A region tree describes the relationship between
%logical regions and partitions.  A region tree contains two types of nodes:
%region nodes and partition nodes.  Every region tree is rooted with a region
%node and alternates between region and partition nodes at each corresponding level.
%A region node represents a specific logical region and tracks the set of
%partitions of that logical region.  Similarly a partition node represents 
%a specific partitioning of a logical region and keeps track of the logical regions 
%which are subregions of the partition.  
\subsubsection{Dependencies From Coherence and Privileges}
\label{subsec:cohdep}
Conceptually, dependence detection between a new task $t$ and a previous task $t'$ requires comparing the two sets of
logical regions accessed.  For each logical region from $t'$
that may overlap with a logical region needed by $t$, the privileges
and coherence modes are compared to determine whether a dependence exists.  If
both regions need read
privileges there is never a dependence, but if either 
task needs write or reduction privileges, the coherence modes are compared:

@@

To detect dependencies efficiently, the high-level runtime maintains a
local region forest (recall Section~\ref{sec:exec}) for each task.  The
root regions are the task's region arguments, and the forest includes
all partitions and subregions created within the task.  To perform
dependence detection on a region $r$, the runtime first finds a root
region $r'$ such that $r \rleq r'$.  The runtime then walks the region
forest from $r'$ to $r$.  Each region $r''$ on the path contains a
list $L$ of registered tasks using $r''$, and potential dependencies
are discovered by comparing privileges and coherence modes with the
tasks in $L$.

The runtime next checks if a partition of $r''$ is {\em open}.  An
open partition indicates that there are active tasks using subregions
of $r''$.  No more than one partition of a node can be open
at any time.  If the open partition is is along the path to $r$,
traversal continues.  If the open partition lies along another 
path the runtime must {\em close} the currently active partition because it cannot guarantee
disjointness with the region the new task wishes to access.  To close a
partition, the runtime finds all active tasks in that
subtree and considers them all dependencies for the new task.  
With the previously active partition closed, the runtime can 
then open the desired partition and traverse further down the tree as
needed.

@@

A {\em physical instance} is a copy of the data in a logical region.  Multiple
physical instances of a logical region may exist simultaneously, and it is the responsibility of the high-level runtime
to ensure tasks use the correct version.
In an out-of-order processor this problem is solved by register-renaming which maps logical
registers to a larger set of physical registers.  Our solution employs a similar technique
that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
is a place holder for the set of physical instances a task will reference when it actually runs.  
Abstract instances are needed because we may not know the actual physical instances a task will
use until its predecessors in the dependence graph are {\em mapped} 
(see Section \ref{subsec:mapping}).  
To determine the abstract instances of a region, we keep track of 
the previous valid abstract instances that were observed while traversing the region forest
during dependence analysis.  Every region in the region forest that has active tasks
has a valid abstract instance.  Every time we close a region we close all the valid
abstract instances.  When a region is reopened it must then create a new abstract instance
to reflect that new physical instances must be created (i.e., a fresh version of the data must be copied from
the parent region).  
%In addition, whenever a task writes a region, a new abstract instance is usually created
%to represent the new version of the data.

@@

For the ghost node regions, dependence analysis discovers 
a dependence between each of the {\tt distribute\_charge} tasks and all of the
{\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
performing writes and the partition is aliased.  However, the runtime will detect these as
anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
create new valid abstract instances rather than use the same abstract instances as
each of the {\tt compute\_new\_current} tasks.  There are no dependences detected
between each of the {\tt distribute\_charge} tasks on the ghost regions because
the {\tt Simultaneous} coherence property permits multiple tasks to be using the same
partition at the same time even if the subregions they are accessing are aliased.  The
state of the region forest after dependence analysis for the first {\tt distribute\_charge} task can
be seen in figure~\ref{sfig:mapping_fig:dc}.

@@

\subsection{Mapping}
\label{subsec:mapping}
%After dependence analysis a task is placed on either a {\em ready queue} to be mapped or a 
The high-level runtime maintains the invariant that a task {\em maps} only after all tasks on which it depends have mapped. 
While some predecessor tasks have not mapped a task remains in the {\em waiting queue}; otherwise it is placed on the {\em ready queue}.

@@

These eight mapper calls give the programmer complete control over where all tasks are executed, and
where all data is placed within the memory hierarchy.  While this interface allows the programmer
to tune performance, it cannot affect correctness.  The high-level runtime still manages all the task dependences and the
ordering in which tasks are executed and copies are performed.  Therefore, regardless of choices
made by any mapper, the resulting program will always be executed correctly.
To make it easy for programmers to write programs, we provide a default
 mapper that uses simple heuristics to place tasks and data.  
%For example, when
%performing task stealing, the default mapper only encourages stealing tasks in groups that use
%the same logical regions, which improves locality by moving tasks that use the same regions
%together.  The default mapper implementation makes it easy for the programmer to get an 
%initial program running.  As the programmer tunes the application, he can write mappers that
%specialize for both application specific and/or machine specific parameters.

The high-level runtime also supports the use of multiple mappers in the same application.  Each task 
call may (but does not need to) have an associated mapper,
%runtime  can choose a mapper to be used for the runtime calls made inside that
%task.  
which allows for the creation of library code that provides its own mappers independent
from any application level mappers.

@@

When a processor notices that it has no work on its ready queue, it will inquire into each
mapper asking for a target processor to attempt a steal.  A mapper can either respond with
a target processor, or indicate that no steal should be attempted.  The runtime will then
send steal requests to each processor on behalf of each mapper.

When a steal request is received at a processor it will determine the mapper(s) that generated
the list of steal requests.  For each mapper that issued a steal request, the runtime will
give the local instance of that mapper a list of the tasks in the ready queue that that
specific mapper is responsible for mapping.  The mapper must then determine the set of tasks which
it is going to permit to be stolen, if any.  Any tasks that the mapper marks as stolen are
sent back to the stealing processor.  By only allowing mappers to make steal requests about
the tasks they own, we give mappers complete control over where tasks are mapped.
In addition, we also give the mappers control over how steal requests are made for
the case where the mapper has better knowledge about where additional work is than the 
runtime.  Note that this mapping interface is sufficiently powerful to allow a user to
disable all stealing if desired, and to map tasks to specific processors with no interference
from the runtime.

\subsection{Execution}
\label{subsec:execution}
Once task mapping is complete the high-level runtime performs the necessary operations
to issue the task to the low level runtime.  It first creates any new physical instances
for the regions the task requires.  It then issues copy operations required
for the task to execute contingent on predecessor tasks producing the source physical instances
having finished.  Then it issues the task to the low level processor contingent
on all the copies for the task having finished as well as the the set of tasks
on which the issuing task depends having finished.  These dependences are all encoded
using {\em events} (see Section \ref{subsec:events}).  Continuing the analogy of an out-of-order
processor, the execution phase is equivalent to issuing an instruction to an
execution unit, and having the execution unit defer the instruction's execution until all its
operands are ready.  The low-level runtime contains processors that act as execution
units that will run a task once all of the events required for the task to run
have occurred.

@@

\subsection{Task Completion}
\label{subsec:cleanup}
We define a task to have {\em completed} once all of the task's regions reflect
the updated state of the task and all its sub-tasks.  
%This reflects an underlying assumption 
%of the mapping process that the regions for which a task has priviliges will represent the data produced 
%by that task.  However, this is not inherently true that a physical instance of a task reflects
%this state after the task finishes executing.  
It is possible for a task to terminate but to have subtasks that are still in the execution phase.
%If a task has child tasks these tasks may
%still be executing when the parent task finishes the execution phase.  
To handle this scenario the runtime must compute the necessary copy operations
from child task physical instances back to the parent task's physical instances.  To
compute these copy operations, the runtime must first wait until all the child tasks
have been mapped.  The runtime then issues a close operation to the each of the parent
task's region forests.  These close operations record all of the necessary copies to
restore the parent task's regions.  
%We call these copy operations at the end of  a task {\em restoring copies}.
The runtime issues these copies with the required dependences
and marks the task completed once the copy operations are complete.
This is similar to the write-back phase of an out-of-order processor where store instructions cannot be
considered complete until the store effects the state of memory.
